Hybrid MILP to efficiently and accuratly \\ solve hard DNN verification instances

### 2024-09-19 ICLR25/formula.tex
section 4

\section*{Formula}

The formula is to choose some nodes and set the corresponding variables to be binary variable in the MILP model. For convenience, we will say that we open those nodes, or nodes are opened to describe that those nodes' variable are set to be binary. In our algorithm, we will open nodes at most 3 layer3 before the target layer. So the formula will consists of three parts.

\subsubsection*{Observation}

The key of our formula is based on the following observation:

\begin{definition}[Improvement]

	Let $a$ be a node in the network (in the second hidden layer or deeper), and we have computed a series of lower and upper bound for all nodes before the layer of $a$. If we consider to compute the upper bound or lower bound of $a$ by the standard MILP model, then

	\begin{enumerate}
		\item Let $M^a_X$ mean the MILP model with target node $a$ and open node set $X$. Hence $M^a_{\emptyset}$ is the LP model for $a$.

		\item For a set $X$, if the upper bound of $a$ computed from $M^a_{\emptyset}$ is $u_0$, and from $M^a_X$ is $u_X$, then the improvement (of upper bound of $a$) of $X$, $I_X^{a,u}$ (or $I_X$ if there is no risk of confusion) is defined as : $$I_X^{a,u} = u_0-u_X.$$ Similarly we can define the improvement of lower bound to be $I_X^{a,l} = l_X-l_0$.
	\end{enumerate}
\end{definition}

Our observation (by experiments) is that \begin{align}
	I_X \approx \sum_{b\in X} I_b.
\end{align} Especially, if all nodes in $X$ are from one layer before $a$, then in experiments, we observe that

\begin{align*}
	|(I_X - \sum_{b\in X} I_b)/I_X| < 1\%. \ (\text{in experiments})
\end{align*} Even $X$ contains nodes from 3 layers before the target layer, in experiments, $I_X$ is still close to $\sum_{b\in X} I_b$.

@@

\subsubsection*{Compute the improvement of a single node}

For one layer before the target layer, the formula is simple and most accurate. Suppose $b$ is a node in one layer before $a$. To estimate $I_b$, first we need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

\begin{definition}
	Let $v(b)$ be the value of $b$ in the LP solution of upper bound of $a$, and $v(\hat{b})$ be the value of $\hat{b}=\ReLU(b)$.	Then the formula to estimate improvement of upper bound of $b$ is: \begin{align}
		I_b \approx W_{ba}(v(\hat{b})-\ReLU(v(b))).
	\end{align}

	Specially, if $v(\hat{b})=\ReLU(v(b))$, then we will have \begin{align*}
		I_b = W_{ba}(v(\hat{b})-\ReLU(v(b))) = 0.
	\end{align*}

>>>>

\section{Important neurons for accuracy and Utility function.}

In this section, we evaluate how each neuron would impact the accuracy of encoding this neuron with an integer (actually binary) or a linear variable. We will say we open a node when this node is selected with a binary variable. We denote by $X$ a choice of set of open nodes, and by $\mathcal{M}_X$ the MILP model where variables from $X$ are encoded with binary variables, and other variables are using the LP linear relaxation.

Let $z$ be a neuron of the DNN we want to evaluate, for instance computing an upper bound on the value it can take. We denote by $n< z$ the fact that a neuron appears on a layer before the layer of $z$, and $n \leq z$ if it can also be $z$.
Let us denote (Sol\_max$_X^z(n))_{n \leq z})$ is a solution of $\mathcal{M}_X$ 
maximizing $z$. In particular, Sol\_max$_X^z(z)$ is the maximum of $z$ under $\mathcal{M}_X$.

Consider (Sol\_max$_\emptyset^z(n))_{n \leq z})$ (this is obtained by {\em one} call to an LP solver, so can be obtained very efficiently).
An accurate utility function is to evaluate 
Improve\_max$^z(a)=$ Sol\_max$_\emptyset^z(z) -$ Sol\_max$_{\{a\}}^z(z)$, 
representing how much opening neuron $a < z$ reduces the maximum computed for $z$
compared with using only LP. 
Similarly, we define (Sol\_min$_\emptyset^z(n))_{n \leq z})$ and 
Improve\_min$_a^z$. Calling MILP on $\mathcal{M}_{\{a\}}$ for each neuron $a \leq z$
would however be very time consuming when the number of neurons $a$ to evaluate is large.

Instead, we focus on the following upper bounding of Improve\_max$_a^z$, thanks to Proposition \ref{LP}.

%\subsubsection*{Observation}

%The key of our formula is based on the following observation:

%Our observation (by experiments) is that \begin{align}
%	I_X \approx \sum_{b\in X} I_b.
%\end{align} Especially, if all nodes in $X$ are from one layer before $a$, then in %experiments, we observe that

\iffalse
\begin{align*}
	|(I_X - \sum_{b\in X} I_b)/I_X| < 1\%. \ (\text{in experiments})
\end{align*} Even $X$ contains nodes from 3 layers before the target layer, in experiments, $I_X$ is still close to $\sum_{b\in X} I_b$.

@@

In our algorithm, we will open nodes at most 3 layer3 before the target layer. So the formula will consists of three parts.

\subsubsection*{Compute the improvement of a single node}
\fi

\subsection*{One Layer before $z$}

%For one layer before the target layer, the formula is simple and most accurate. 
Suppose $a$ is a node in one layer before $z$. To estimate Improve\_max$_a^z$:
%we first need to run $M^a_{\emptyset}$ to compute the upper bound of $a$  to obtain the solution data. Especially, we will read the values of $b$, before $\ReLU$ function and after $\ReLU$ function.

\begin{definition}
	For all nodes $n$, let $sol(n)=$Sol\_max$_\emptyset^z(n)$ be the value of node $a$
	in the solution of the LP instance Sol\_max$_\emptyset^z$ to maximize $z$.

	We define : 
	\begin{align}
		Utility\_max^z(a) = W_{az} \times (sol(\hat{a})- \ReLU(sol(a))).
	\end{align}

	%In particular, if $v(\hat{a})=\ReLU(v(a))$, then we will have 
	%\begin{align*}
%		Utility^z(a) = 0.
%	\end{align*}

We have that :

\begin{proposition}
$0 \leq Improve\_max^z(a) \leq Utility\_\max^z(a)$. 
\end{proposition}

\begin{proof}
Indeed, consider $(sol'(n))_{n \leq z})_{n \leq z})$ with
$sol'(n)=sol(n)$ for all $n \notin \{z,\hat{a}\}$.
In particular,  $sol'(a) = sol(a)$.

Now, define $sol'(\hat{a}) = \ReLU(v(a))$. 
That is,  
sol'$(\hat{a})$ is the correct value for $\hat{a}$ (obtained if we open node $a$), 
compared to the LP abstraction for $sol(\hat{a})$.

Last, we define $sol'(z)=sol(z) + W_{az} \times (\ReLU(sol(a)) - sol(\hat{a})) = sol(z) - Utility\_max_z(a)$.

Now, it is easy to check that $(sol'(n))_{n \leq z}$ satisfies the constraints in 
$\mathcal{M}_{\{a\}}$, as openning $a$ changes the value of $\hat{a}$ from
$sol(\hat{a})$ to $\ReLU(sol(a))$, and the contribution from $a$ to $z$ is 
$W_{az}$. As $sol'(z)$ is a solution of $\mathcal{M}_{\{a\}}$, it is smaller or equal to the maximal solution: $sol'(z) \leq$ Sol\_max$_{\{a\}}^z(z)$. That is, 
$sol(z)-sol'(z) \geq sol(z) -$ Sol\_max$_{\{a\}}^z(z)$, i.e. 
$Improve\_max^z(a) \leq Utility\_\max^z(a)$.
\end{proof}
